---
title: "repoRter.nih: a convenient R interface to the NIH RePORTER Project API"
author:
- Michael Barr, ACAS, MAAA, CPCU
output: 
  pdf_document:
    keep_md: yes
    keep_tex: yes
    pandoc_args: --listings
  html_document:
    highlight: tango
    theme: flatly
header-includes:
- \UseRawInputEncoding
- \usepackage{xcolor}
- \lstset{ backgroundcolor=\color[RGB]{248,248,248}, basewidth=0.5em, basicstyle=\small\ttfamily,
  breakatwhitespace=true, breakautoindent=true, breaklines=true, captionpos=b, columns=flexible,
  commentstyle=\color[rgb]{0.56,0.35,0.01}\itshape, escapeinside={\%*}{*)}, frame=tb,
  keepspaces=true, keywordstyle=\color[rgb]{0.13,0.29,0.53}\bfseries, linewidth=\textwidth,
  numberstyle=\footnotesize, showspaces=false, showstringspaces=false, showtabs=false,
  stringstyle=\color[rgb]{0.31,0.60,0.02}, tabsize=2, }
vignette: >
  %\VignetteIndexEntry{repoRter.nih: a convenient R interface to the NIH RePORTER Project API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  width = 95,
  fig.dpi = 96,
  cache.path = "cache/"
)

options(width = 95, cli.unicode = FALSE, cli.width = 95)
```

```{r, echo = FALSE, message = FALSE, fig.width=6, fig.height=4, fig.align='center'}
library(tibble)
library(repoRter.nih)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(scales)
library(tufte)
data("covid_response_codes")

req <- make_req(criteria =
                  list(covid_response = c("All")),
                include_fields = nih_fields %>%
                  filter(payload_name %in% c("award_amount_range", "covid_response"))
                %>% pull(include_name))

res <- get_nih_data(req,
                    flatten_result = TRUE)
message("plot 1")

p <- res %>%
  left_join(covid_response_codes, by = "covid_response") %>%
  mutate(covid_code_desc = case_when(!is.na(fund_src) ~ paste0(covid_response, ": ", fund_src),
                                     TRUE ~ paste0(covid_response, " (Multiple)"))) %>%
  group_by(covid_code_desc) %>%
  summarise(total_awards = sum(award_amount) / 1e6) %>%
  ungroup() %>%
  arrange(desc(covid_code_desc)) %>%
  mutate(prop = total_awards / sum(total_awards),
         csum = cumsum(prop),
         ypos = csum - prop/2 ) %>%
  ggplot(aes(x = "", y = prop, fill = covid_code_desc)) +
  geom_bar(stat="identity") +
  geom_text_repel(aes(label =
                        paste0(dollar(total_awards,
                                      accuracy = 1,
                                      suffix = "M"),
                               "\n", percent(prop, accuracy = .01)),
                      y = ypos),
                  show.legend = FALSE,
                  nudge_x = .8) +
  coord_polar(theta ="y") +
  theme_void() +
  theme(legend.position = "right",
        legend.text = element_text(colour="blue", size=6, 
                                   face="bold")) +
  ggtitle("Legislative Source for NIH Covid Response Project Funding")

rm(req, res)
detach("package:repoRter.nih", unload = TRUE)
message("past the plot")
p
```

# Introduction

The US National Institute of Health (NIH) received approximately \$42 billion in funding fiscal year 2022; \$31 billion (72%) of this was awarded by the NIH in the form of grant funding to hospitals, medical colleges, non-profits, businesses, and other organizations based in the U.S. and abroad.^[https://nexus.od.nih.gov/all/2021/04/21/fy-2020-by-the-numbers-extramural-investments-in-research] The NIH maintains a publicly available database called "RePORTER" to track this substantial flow of grant funding and makes it available to the public via [a web-based query interface](https://reporter.nih.gov/) as well as an API.

> "The NIH RePORTER APIs is designed to programmatically expose relevant scientific awards data from both NIH and non-NIH federal agencies for the consumption of project teams or external 3rd party applications to support reporting, data analysis, data integration or to satisfy other business needs as deemed pertinent." 
>`r tufte::quote_footer('--NIH RePORTER v2 API Documentation')`

This data data can have significant value for researchers, investors, industry, watchdogs/public advocates, and R users, to name a few But constructing queries and retrieving results programmatically involves some coding overhead which can be a challenge for those not familiar with RESTful APIs and JSON; it takes some effort even for those who are. The `repoRter.nih` package aims to simplify this task for the typical analyst scripting in R.

# Getting Started

## Installation

This package (latest stable release) can be installed from CRAN the usual way:
```{r, eval = FALSE}
install.packages("repoRter.nih")
```

The current dev version can be installed from github, on the `dev` branch:
```{r, eval = FALSE}
devtools::install_github('bikeactuary/repoRter.nih@dev')
```

I welcome more capable R developers than myself to collaborate on improving the source code, documentation, and unit testing in this package.

# Basic Workflow

```{r, message = FALSE}
library(repoRter.nih)
```

The `make_req()` method is used to generate a valid JSON request object. The req can subsequently be passed to the RePORTER Project API and results retrieved via the `get_nih_data()` method.

Generating the request:
```{r}
# all projects funded by the Paycheck Protection Act, Coronavirus Response and
# Relief Act, and American Rescue Plan, in fiscal year 2021
req <- make_req(criteria =
                  list(fiscal_years = 2021,
                       covid_response = c("C4", "C5", "C6")))
```

Sending the request and retrieving results

```{r, cache = TRUE}
res <- get_nih_data(req)
class(res)
```

A tibble is returned containing 43 columns, several of them nested `data.frame`s and lists (of variable length vectors and `data.frame`s of varying height).

```{r}
res %>% glimpse(width = getOption("cli.width"))
```

# Criteria-Field Translation

A dataset (`nih_fields`) is provided with this package to assist in translating between field names used in the payload `criteria`, column names in the return data, and field names used in the `include_fields`, `exclude_fields`, and `sort_field` arguments.

```{r}
data("nih_fields")
nih_fields %>% print
```
Some fields can not be used as filtering `criteria` - these will show `NA` in the `payload_name` column.

# Generating Requests

Most of the detail (and function documentation) is around the many parameters available in RePORTER to filter/search project records. Let's get into some of the capabilities.

## Default Request

If no arguments are supplied, the default behavior of `make_req()` is to generate a request for all projects funded in `fiscal_years = lubridate::year(Sys.Date())`. Limiting requests to a single year will often be necessary due to a RePORTER restriction that a maximum of 100K records may be returned from any result set - there are currently ~2.6M projects in the database going back to fiscal year 1985.

```{r}
req <- make_req()
```

The method prints a helpful message to the console in addition to returning the JSON. Set `message = FALSE` if you wish to suppress this message.

## Limiting data retrieved

You can limit both the width and height of the result set retrieved from the API.

### Fields

We probably will not need to fetch every field every time. The `include_fields` argument is provided to specify a limited set of fields to be returned. Alternatively, fields may be excluded using `exclude_fields`. 

### Records (projects)

This package provides the ability to retrieve just a limited number of result pages via the `max_pages` argument. This can be useful for developing/testing your queries (and for reducing time to render package documentation). Each page has a record count equal to `limit` - so setting `max_pages = 5` with the default `limit = 500` in `make_req()` will result in up to 2,500 total records returned.

### Example 1 - limiting results and selecting fields

```{r}
data("nih_fields")
fields <- nih_fields %>%
  filter(response_name %in% 
           c("appl_id", "subproject_id", "project_title", "fiscal_year",
             "award_amount", "is_active", "project_start_date")) %>%
  pull(include_name)

req <- make_req(include_fields = fields,
                limit = 500,
                message = FALSE) # default
res <- get_nih_data(query = req,
                    max_pages = 1)

res %>% glimpse(width = getOption("cli.width"))
```

## Some Vanilla Criteria

Many criteria are passed as vectors within the `criteria` list argument. We will cover some of the most useful examples.

### Ex. 2 - Organization search

We can refine our query results by providing filtering criteria to `make_req()`, and by extension to the API. Suppose we want all currently active projects, funded in fiscal years 2017 through 2021, with a specific organization in mind (though we don't know exactly how its name will appear in RePORTER):

```{r}
req <- make_req(criteria = 
                  list(
                    fiscal_years = 2010:2011,
                    include_active_projects = TRUE, # LIMITS to active projects
                    org_names = c("Yale", "New Haven")
                  ),
                include_fields = c("Organization", "FiscalYear", "AwardAmount"),
                message = FALSE)
```

Here we are asking for any orgs containing the strings "yale" or "new haven" (ignoring case) - there are implied wildcards on either end of the strings we provide. This is the same as `org_name LIKE '%yale%' OR org_name LIKE '%new haven%'" in a SQL WHERE clause.

```{r, cache=TRUE}
res <- get_nih_data(req, max_pages = 1)
res %>% glimpse(width = getOption("cli.width"))
```

Look closely at the print output and notice the column `organization` is a nested data frame - it has 17 columns and always a single record. Setting `flatten_result = TRUE` in the call to `get_nih_data()` will flatten all such return fields, prefixing the original field name and returning with clean names (see `janitor::clean_names`).

```{r}
res <- get_nih_data(req,
                    max_pages = 1,
                    flatten_result = TRUE)

res %>% glimpse(width = getOption("cli.width"))
```

Most users will prefer the flattened format above. It looks like Yale is busy, but it is not the only org matching our search.

```{r}
res %>% 
  group_by(organization_org_name) %>%
  summarise(project_count = n())
```

The `org_names_exact_match` criteria can be used as an alternative when we know the exact org name as it appears in RePORTER, if we want only that org's projects returned.

### Ex. 3 - Geographic search

We can also filter projects by the geographic location (country/state/city) of the applicant organization.

```{r, cache=TRUE}
## A valid request but probably not what we want
req <- make_req(criteria = 
                  list(
                    fiscal_years = 2010:2011,
                    include_active_projects = TRUE,
                    org_cities = "New Haven",
                    org_states = "WY"
                  ),
                include_fields = c("Organization", "FiscalYear", "AwardAmount"),
                message = FALSE ## suppress printed message
)

res <- get_nih_data(req,
                    max_pages = 5,
                    flatten_result = TRUE)
```

Multiple criteria are usually connected by logical "AND" - there are no orgs based in the city of New Haven in Wyoming state (because it doesn't exist.)

```{r}
req <- make_req(criteria =
                  list(
                    fiscal_years = 2015,
                    include_active_projects = TRUE,
                    org_states = "WY"
                  ),
                include_fields = c("ApplId", "Organization", "FiscalYear", "AwardAmount"),
                sort_field = "AwardAmount",
                sort_order = "desc",
                message = FALSE)

res <- get_nih_data(req,
                    flatten_result = TRUE)

res %>% glimpse(width = getOption("cli.width"))
```

Why are there projects from more recent years than 2015? Because the `include_active_projects` flag adds in active projects that match all criteria aside from `fiscal_years` (this appears to be the intended behavior by RePORTER).

### Ex. 3 - Coronavirus/Covid-19 research

We already provided one example of this search criteria above. Let's mix it up and request all covid response projects.

```{r, cache=TRUE}
## all projects funded by the Paycheck Protection Act, Coronavirus Response and Relief Act,
## and American Rescue Plan, in fiscal year 2021
req <- make_req(criteria =
                  list(covid_response = c("All")),
                include_fields = nih_fields %>%
                  filter(payload_name %in% c("award_amount_range", "covid_response"))
                %>% pull(include_name))

res <- get_nih_data(req)
res$covid_response %>% class()
res$covid_response[[1]]
```

`covid_response` is a nested list (with character vectors of variable length) within the return tibble. We can use `flatten_result = TRUE` here - elements of each vector will be collapsed to a single string with ";" as separator, converting the list to a single character vector.

```{r covid, cache = TRUE}
## all projects funded by the Paycheck Protection Act, Coronavirus Response and Relief Act,
## and American Rescue Plan, in fiscal year 2021
req <- make_req(criteria =
                  list(covid_response = c("All")))

res <- get_nih_data(req,
                    flatten_result = TRUE)

unique(res$covid_response)
```

Summarizing all Covid-related project awards:

```{r, fig.width=6, fig.height=4}
library(ggplot2)

message("plot 2")

res %>%
  left_join(covid_response_codes, by = "covid_response") %>%
  mutate(covid_code_desc = case_when(!is.na(fund_src) ~ paste0(covid_response, ": ", fund_src),
                                     TRUE ~ paste0(covid_response, " (Multiple)"))) %>%
  group_by(covid_code_desc) %>%
  summarise(total_awards = sum(award_amount) / 1e6) %>%
  ungroup() %>%
  arrange(desc(covid_code_desc)) %>%
  mutate(prop = total_awards / sum(total_awards),
         csum = cumsum(prop),
         ypos = csum - prop/2 ) %>%
  ggplot(aes(x = "", y = prop, fill = covid_code_desc)) +
  geom_bar(stat="identity") +
  geom_text_repel(aes(label =
                        paste0(dollar(total_awards,
                                      accuracy = 1,
                                      suffix = "M"),
                               "\n", percent(prop, accuracy = .01)),
                      y = ypos),
                  show.legend = FALSE,
                  nudge_x = .8) +
  coord_polar(theta ="y") +
  theme_void() +
  theme(legend.position = "right",
        legend.text = element_text(colour="blue", size=6, 
                                   face="bold")) +
  ggtitle("Legislative Source for NIH Covid Response Project Funding")
```

A second dataset is provided to translate the `covid_response` codes:

```{r}
data("covid_response_codes")
covid_response_codes %>% print
```

# Some Rocky Road Criteria

Other criteria provide search and filtering capability on many of the nested data elements. These criteria are passed as lists and must include a value for each of the named elements within.

### Example 4 - Principal Investigator/Officer name search

The `pi_names` and `po_names` criteria allow the user to search for projects based on the first and last names of Principal Investigators and Principal Officers assigned. Each of these criteria must be provided as a list with 3 named character vector elements: `first_name`, `last_name`, and `any_name`. _Even if you only want to search on one of these name fields, you must provide the remaining elements as an empty string._ We will demonstrate with a search on PI name.

```{r}
## projects funded in 2021 where the principal investigator first name
##  is "Michael" or begins with "Jo"
req <- make_req(criteria = 
                  list(fiscal_years = 2021,
                       pi_names = list(first_name = c("Michael", "Jo*"),
                                       last_name = c(""), # must specify all pi_names elements always
                                       any_name = character(1))),
                include_fields = nih_fields %>%
                  filter(payload_name == "pi_names") %>%
                  pull(include_name),
                message = FALSE)

res <- get_nih_data(req,
                    max_pages = 1,
                    flatten_result = TRUE)

res %>% glimpse(width = getOption("cli.width"))
```

Here we searched for any projects with a PI first-named "Michael" or beginning with "Jo" - the "*" is a wildcard operator.

Note that the first column in the return is a list of data frames of _variable height_ - we leave such returned elements to the user to handle extraction - flattening is only done for lists of atomic vectors and nested data frames.

### Ex. 5 - Advanced text search

RePORTER allows users to search the project title, abstract, and tags for specific terms or phrases. You can access this capability with the `advanced_text_search` criteria - a named list with 3 elements.

- `operator` may be either "and", "or", or "advanced" - and/or will specify the logical operator connecting multiple search terms. "advanced" allows the user to pass a boolean search string directly.
- `search_field` can be any or multiple of "terms", "abstract", "projecttitle." To search all items, specify "all" or "" (a length 1 vector with an empty string).

```{r}
## using advanced_text_search with boolean search string
req <- make_req(criteria = 
                  list(advanced_text_search =
                         list(operator = "advanced",
                              search_field = c("terms", "abstract"),
                              search_text = "(head AND trauma) OR \"brain damage\" AND NOT \"psychological\"")),
                include_fields = c("ProjectTitle", "AbstractText", "Terms")  )

res <- get_nih_data(req, max_pages = 1)
```
Let's read the searched fields for one of these records
```{r}
one_rec <- res %>%
  slice(4) %>%
  mutate(abstract_text = gsub("[\r\n]", " ", abstract_text))

one_rec %>% pull(project_title) %>% print
```

```{r}
one_rec %>% pull(abstract_text) %>% print
```

```{r}
one_rec %>% pull(terms) %>% substr(1, 500) %>% print
```


# Additional Resources

- In addition to this vignette and function documentation, the [RePORTER web interface](https://reporter.nih.gov/advanced-search) is another useful resource for getting familiar with available search parameters

- The official API documentation is located [here](https://api.reporter.nih.gov/documents/Data%20Elements%20for%20RePORTER%20Project%20API%20v2.pdf)

- ...and the homepage with further examples/documentation is [here](https://api.reporter.nih.gov/)

- Information on NIH study sections, IRGs, etc. is [here](https://public.csr.nih.gov/StudySections)

- h/t to Chris whose [code on github](https://github.com/christopherBelter/nih_reporter_api) I found and served as a starting point for this work

# Session Information

The version number of R and packages loaded for generating the vignette were:

```{r}
sessionInfo()
```
